---
title: "Exercise 2"
author: "Claudia Dorigo"
date: "01/04/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**DAAG: EXERCISES chapter 3 (11,13); chapter 4 (6,7)**
 
***11***

The following data represent the total number of aberrant crypt foci (abnormal growths in the colon) observed in seven rats that had been administered a single dose of the carcinogen azoxymethane and sacrificed after six weeks (thanks to Ranjana Bird, Faculty of Human Ecology, University of Manitoba for the use of these data):
87 53 72 90 78 85 83
Enter these data and compute their sample mean and variance. Is the Poisson model appropriate
for these data? To investigate how the sample variance and sample mean differ under the Poisson
assumption, repeat the following simulation experiment several times:
x <- rpois(7, 78.3)
mean(x); var(x)

***Solution: ***
```{r}
data <- c(87, 53, 72, 90, 78, 85, 83)


rep <- 1000
x <- array(0,c(rep,2))

for (i in 1:rep){
  xx <- rpois(7, 78.3)
  x[i,1] <- mean(xx)
  x[i,2] <- var(xx)
}

# plot histograms of simulated sample mean and sample variance
# plot also a red line for the observed values
par(mfrow=c(1,2))
hist(x[,1],breaks=20,main="sample mean of poisson")
abline(v=mean(data),col=2)
hist(x[,2],breaks=20,main="sample variance of poisson")
abline(v=var(data),col=2)

#prob of getting a sample variance bigger than the observed one
sum(x[,2]>var(data))/rep

# compute the p-value from the theoretical distribution
# because I know the distribution of the variance:
#(n-1)*s2/sigma^2 ~ chi(n-1) where sigma^2 is lambda for a poisson
chi_obs <- var(data)*6/78.3
1-pchisq(chi_obs,6)
#small but I can't reject the hypotesis
```


***13***

 A Markov chain for the weather in a particular season of the year has the transition matrix, from
one day to the next:
P b =
    Sun Cloud Rain
Sun 0.6 0.2 0.2
Cloud 0.2 0.4 0.4
Rain 0.4 0.3 0.3

It can be shown, using linear algebra, that in the long run this Markov chain will visit the states
according to the stationary distribution:
Sun Cloud Rain
0.641 0.208 0.151
A result called the ergodic theorem allows us to estimate this distribution by simulating the
Markov chain for a long enough time.

(a) Simulate 1000 values, and calculate the proportion of times the chain visits each of the states. Compare the proportions given by the simulation with the above theoretical proportions.

(b) Here is code that calculates rolling averages of the proportions over a number of simulations and plots the result. It uses the function rollmean() from the zoo package.
```{r}
plotmarkov <-
function(n=10000, start=0, window=100, transition=Pb, npanels=5){
xc2 <- Markov(n, start, transition)
mav0 <- rollmean(as.integer(xc2==0), window)
mav1 <- rollmean(as.integer(xc2==0), window)
npanel <- cut(1:length(mav0), breaks=seq(from=1, to=length(mav0),
length=npanels+1), include.lowest=TRUE)
df <- data.frame(av0=mav0, av1=mav1, x=1:length(mav0),
gp=npanel)
print(xyplot(av0+av1~x | gp, data=df, layout=c(1,npanels),
type="l", par.strip.text=list(cex=0.65),
scales=list(x=list(relation="free"))))
}
```
Try varying the number of simulations and the width of the window. How wide a window
is needed to get a good sense of the stationary distribution? This series settles down rather
quickly to its stationary distribution (it “burns in” quite quickly). A reasonable width of
window is, however, needed to give an accurate indication of the stationary distribution


***6***
Here we generate random normal numbers with a sequential dependence structure:
```{r}
y1 <- rnorm(51)
y <- y1[-1] + y1[-51]
acf(y1) # acf is ‘autocorrelation function’
# (see Chapter 9)
acf(y)
```

Repeat this several times. There should be no consistent pattern in the acf plot for different
random samples y1. There will be a fairly consistent pattern in the acf plot for y, a result of
the correlation that is introduced by adding to each value the next value in the sequence.

***Solution: ***
```{r}
par(mfrow=c(4,4))

for (i in 1:8){
  y1 <- rnorm(51)
y <- y1[-1] + y1[-51]
acf(y1)
acf(y)
}
```
As expected I can see that for Series y the autocorrelation function is not significantly different from 0 for any lag which is not 0 (the function is below the significance level represented by the dashed line), while of course the autocorrelation function when the lag is equal to 0 is 1 because it's the autocorrelation between each observation and itself. This tells us that a sample generated using $\textit{rnorm}$ function is composed by independent observations.

In the plots of the autocorrelation function for Series y1 instead we can see that in each simulation the values corresponding to a lag equal to 1 is always significantly different from 0 (and also when lag=0 for the same reason of the Series y). This is an expected result because of how I've built the Series y, indeed I've build each observation depending on itself and the previous one.

***7***
Create a function that does the calculations in the first two lines of the previous exercise.
Put the calculation in a loop that repeats 25 times. Calculate the mean and variance for each vector y that is returned. Store the 25 means in the vector av, and store the 25 variances in the vector v. Calculate the variance of av. 

**CORE STATISTICS chapter 3 (3.3, 3.5)**

**3.3** 
Rewrite the following, replacing the loop with efficient code:
```{r}
n <- 100000
z <- rnorm(n)
zneg <- 0
j <- 1
for (i in 1:n) {
if (z[i]<0) {
zneg[j] <- z[i]
j <- j + 1
}
}
```
Confirm that your rewrite is faster but gives the same result.

### Solution:
```{r}
n <- 100000
z <- rnorm(n)
zneg <- 0
zneg2 <- 0
j <- 1

# timing the given function
start <- Sys.time()
for (i in 1:n) {
  if (z[i]<0) {
    zneg[j] <- z[i]
    j <- j + 1
  }
}
end <- Sys.time()
time1 <- end-start

# timing my optimized function
start <- Sys.time()
zneg2=z[z<0]
end <- Sys.time()
time2 <- end-start

# check which one is faster
time1
time2 #faster

#chech if the results are the same:
summary(zneg)
summary(zneg2)

```


**3.5**
 Consider solving the matrix equation $Ax = y$ for x, where y is a known $n-$
vector and A is a known n×n matrix. The formal solution to the problem is
$x = A^{−1}y$, but it is possible to solve the equation directly, without actually
forming $A^{−1}$. This question explores this direct solution. Read the help file
for $\textit{solve}$ before trying it.

a. First create an A, x and y satisfying Ax = y.
```{r}
set.seed(0)
n <- 1000
A <- matrix(runif(n*n),n,n)
x.true <- runif(n)
y <- A%*%x.true
```
The idea is to experiment with solving $Ax = y$ for x, but with a known
truth to compare the answer to.

b. Using $\textit{solve}$, form the matrix $A^{−1}$ explicitly and then form $x_1=A^{−1}y$. Note how long this takes. Also assess the mean absolute difference between x1 and x.true (the approximate mean absolute ‘error’ in the solution).
c. Now use solve to directly solve for x without forming $A^{−1}$. Note how
long this takes and assess the mean absolute error of the result.
d. What do you conclude?

### Solution:
b.
```{r}
# here I compute the inverse of A using solve(A) and then I compute x1
start <- Sys.time()
A_inverse <- solve(A)
x1 <- A_inverse%*%y
end <- Sys.time()
time1 <- end-start
```

c.
```{r}
# here I directly compute x1 using solve
start <- Sys.time()
x1 <- solve(A,y) 
end <- Sys.time()
time2 <- end-start
```

d.
check which one is faster
```{r}
# in previous code I've timed both versions
time1
time2 #faster
```


**LAB EXERCISES **

***Exercise 1***
Check the biased nature of s2b via MC simulation, generating n=10 iid values from a normal distribution. Plot also s2 and comment the difference.

***Solution***
```{r}
rep <- 1000
n <- 5
mu <- 5
sigma <- 3

samples <- array(0, c(rep,n))
statistics <- array(0,c(rep,2))


for (i in 1:rep){
  samples[i,] <- rnorm(n,mu,sigma)
  statistics[i,1] <- var(samples[i,])
  statistics[i,2] <- statistics[i,1]*(n-1)/n 
}

#is s2 correct?
mean(statistics[,1])
sigma**2

#is sb2 correct?
mean(statistics[,2])
sigma**2

#plot s2 sb2 and true value
p1 <- hist(statistics[,1],breaks=100)                    
p2 <- hist(statistics[,2],breaks=100)                    
plot(p2,main="biasness of S2 and Sb2")  
plot(p1, col=rgb(1,0,0,1/4), add=T)  
abline(v=sigma**2,lwd=3)  
abline(v=mean(statistics[,1]),col=2,lty=2,lwd=3)
abline(v=mean(statistics[,2]),lty=2,lwd=3)

```



***Exercise 2***
What happens if a great player decides to join you, now? Try to simulate the data and perform the test again.

***Solution:***
I suppose 6 people have prob1 and the great palyer has prob2.
I perform the simulation and I compute the statistic
```{r}
n <- 50 #sample size
K <- 4 #number of zones

# 6 players + 1 great player
prob1 <- c(7/16,5/16,3/16,1/16)
prob2 <- c(5/16, 3/16, 6/16, 2/16)

y <- array(0,c(7,n))
observed <- array(0,c(7,4))
for (i in 1:6){
  y[i,] <- sample( 1:K, n, replace=TRUE, prob=prob1)
}
y[7,] <- sample( 1:K, n, replace=TRUE, prob=prob2)


observed <- apply(y,1,table)
observed
  expected <- c( n*(7/16), n*(5/16), n*(3/16), n*(1/16))

x2 <- sum(apply(((observed-expected)^(2)/expected),1,sum))
pchisq(x2, df =(K-1)*(6), lower.tail =FALSE )
```
I reject the null hypotesis (all players have the same probability) but I don't know how many (or which ones) players have different probabilities.

***Exercise 3***
Sometimes it could be useful to assess the degree of association, or correlation, between paired samples, using the Pearson, the Kendall’s τ or the Spearman’s ρ correlation coefficient. Regardless of the adopted cofficient, the null hypothesis for a given correlation coefficent ρ is:
$$
H_0:ρ=0.
$$
The test statistic is then defined as
$$
T=r\sqrt{\frac{n-2}{1-r^2}}\sim^{H_0} t_{n-2}
$$
where $r=Corr(X,Y)$ is the Pearson correlation coefficient. Suppose to have two samples of the same length $x_1,…,x_n,y_1,…,y_n$, and to measure the association between them. Once we compute the test statistic tobs, we may then compute the p-value (here we are evaluating a two sided test) as:
$$
p=2Pr_{H_0}(T≥|t_{obs}|).
$$
Consider now some of the most followed Instagram accounts in 2018: for each of the owners, we report also the number of Twitter followers (in milions). Are the Instagram and Twitter account somehow associated? Perform a correlation test, compute the p-value and give an answer. Here is the dataframe.
```{r echo=T, results='hide'}
 Owners <- c( "Katy Perry", "Justin Bieber", "Taylor Swift", "Cristiano Ronaldo",
                   "Kim Kardashian", "Ariana Grande", "Selena Gomez", "Demi Lovato")
      Instagram <- c( 69, 98,107, 123, 110, 118, 135, 67)
      Twitter <- c( 109, 106, 86, 72, 59, 57, 56, 56)
      plot( Instagram, Twitter, pch=21, bg=2, xlim=c(60, 150), ylim=c(40, 120) )
      text( Instagram[-6], Twitter[-6]+5, Owners[-6], cex=0.8 )
      text( Instagram[6], Twitter[6]-5, Owners[6], cex=0.8 )
```
I've just plotted the distribution, t_observed and boundaries of acceptance/rejection region
```{r}
Instagram <- c( 69, 98,107, 123, 110, 118, 135, 67)
Twitter <- c( 109, 106, 86, 72, 59, 57, 56, 56)

n=length(Instagram)
r=cor(Instagram,Twitter)
r
t_obs=r*sqrt((n-2)/(1-r^2))

x <- seq(-5,5,length=1000)
plot(x,dt(x,n-2),type="l")
abline(v=t_obs,col=2)

#I fix alpha=0.05 and plot rejection and acceptance region
alpha <- 0.05
r1 <- qt(alpha/2,n-2)
r2 <- qt(1-alpha/2,n-2)
abline(v=r1,lty=2)
abline(v=r2,lty=2)
```


***Exercise 4***
Compute analitically $J(\gamma,\gamma;y),J(\gamma,\beta;y),J(\beta,\beta;y)$.

So we assume $y_i \sim Weibull(\gamma,\beta)$. 
This is the weibull distribution:
$$
p(x|\gamma,\beta)=\frac{\beta}{\gamma^{\beta}}x^{\beta-1}e^-(\frac{x}{\gamma})^\beta
$$
and I have to compute the second derivatives
...

***Exercise 5***
Produce the contour plot for the quadratic approximation of the log-likelihood, based on the Taylor series:
$$
l(\theta)-l(\hat{\theta})≈-\frac{1}{2}(\theta-\hat{\theta})^TJ(\hat{\theta})(\theta-\hat{\theta})
$$
 I have the matrix $\hat{J}$ from previous exercise and the likelihood function. I use the ML estimate for $\hat{\theta}$ from the notebook 
 $$
 l(\theta)≈l(\hat{\theta})-\frac{1}{2}(\theta-\hat{\theta})^TJ(\hat{\theta})(\theta-\hat{\theta})
 $$

